---
title: "Meetup 2 - Importing and Cleaning Data"
author: "Kampalr"
date: "February 24, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = FALSE)
```

For this meetup, we shall be importing and cleaning data obtained from data.ug. We shall be looking at the [PLE Results by district 2010 - 2016 dataset](http://catalog.data.ug/dataset/uneb-results/resource/6e5885bd-2952-4086-aaac-0c2aa58ec762).

We shall levearge on the use of [b]base R functions[/b] as these functions tend to be the most popular for new entrants into R. For our next session, we shall introduce members to [tidyverse](https://www.tidyverse.org/) that provides a collection of functions purposely customised for data science tasks.

### Importing the data
The file that we are using is a standard CSV file containing details on PLE results for the period 2010 - 2016.

```{r dataImport}
url <- "http://catalog.data.ug/dataset/1242faeb-b5a0-4ffb-8274-9c9975aad183/resource/6e5885bd-2952-4086-aaac-0c2aa58ec762/download/ple-results-by-district-2010-2016.csv"

data <- read.csv(url, stringsAsFactors = FALSE)
head(data)
```

The data looks fairly clean from the initial on-set but let's investigate further...

## Cleaning the data
We won't attempt to completely clean the data set as this may take too much time but instead, we shall look at possible techniques to look at the data and invoke thinking on what approach could be taken to clean the dataset.

Let's first look at the spread of the data.

### 1. How many districts are there?

```{r districts}
districts <- readLines(file('districts.txt'))
districts

```

Looking at the [census data (2017)](http://www.ubos.org/onlinefiles/uploads/ubos/census_projections/District%20Single%20Years%20Final_17.07.2017.xls) from Uganda Bureau of Statistics, we can see that there are currently **```r length(districts)```** districts in Uganda.

What happens when we try to count the number of districts that are present in the PLE dataset?  
```{r districtCount}
districts_PLE <- as.data.frame(table(data$DISTRICT))
districts_PLE
```

We can observe that there are ***`r nrow(districts_PLE)`*** unique districts registered. By scrolling through the few rows, we notice a couple of things:

 - There are duplicate districts (e.g. Abim, Adjumani, etc.). The districts have white spaces which prevent them from being recognised as the same.
 - There are subcounties/municipalities listed as districts, e.g. Arua M/C, Arua Main, and Arua Mun.
 
 Let's address the first issue and remove the white space and see whether our results will improve.
 
### Removing white space
 
```{r removeWhiteSpace}
data$DISTRICT <- unlist(lapply(X = data$DISTRICT, FUN = trimws))

# And let's test the new results
districts_PLE <- as.data.frame(table(data$DISTRICT))
districts_PLE

```

As we can see, the duplicate districts have been knocked out bringing the unique number of districts down to ***`r nrow(districts_PLE)`***. Brilliant.

We can now tackle the second observation.

### Standardise districts
We noted that there were a number of districts that were broken down into counties, subcounties and municipalities. As an example, we have Arua district:
```{r Arua} 
unique(data[grep(pattern = "*ARUA*", x = data$DISTRICT, ignore.case = TRUE), 'DISTRICT'])
```

This presents a challenge 